D3064R0: Bringing the Laws of Physics, Computer-System Constraints, and Compilers to Bear on the OOTA Problem

### 2024-02-14 ootavacuous.tex
abstract: Pull in quasi-volatile objects
Note that we only conjecture that reasonable implementations should
adhere to quasi-volatile behavior for relaxed non-volatile atomics.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
introduction: Add roadmap
And a LaTeX comment for a future "Future Directions" section.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

	This paper refactors the commonly used condition for classifying
	OOTA behavior, which permits bringing to bear constraints from
	long-standing laws of physics and from hardware systems.
	In this initial work, we focus on C++ implementations based
	on compilers and computing hardware (including CPUs and GPGPUs).
	We present informal arguments that these constraints and
	limitations prevent OOTA cycles containing only volatile atomics
	from occurring in undefined-behavior-free C++ programs running on
	correctly constructed C++ compiler-based implementations that are
	restricted to single-threaded analysis in our current universe.
	We further conjecture that, given current optimizations, these
	constraints also prevent OOTA cycles involving non-volatile
	atomics.
\end{abstract}

@@

@@@ Add brief description and roadmap of paper, including the simplifications
and restrictions.

>>>>

	This paper refactors the commonly used condition for classifying
	OOTA behavior, which permits bringing to bear constraints from
	long-standing laws of physics and from hardware systems.
	In this initial work, we focus on C++ implementations based
	on compilers and computing hardware (including CPUs and GPGPUs).
	We present informal arguments that these constraints and
	limitations prevent OOTA cycles containing only volatile atomics
	from occurring in undefined-behavior-free C++ programs running on
	correctly constructed C++ compiler-based implementations that are
	restricted to single-threaded analysis in our current universe.
	We extend this work to non-volatile atomics by defining
	quasi-volatile behavior that we conjecture that usable real-world
	C++ implementations must adhere to.
\end{abstract}

@@

This paper shows that C++ implementations subject to reasonable
real-world constraints cannot exhibit OOTA cycles.
Section~\ref{sec:Background}
provides background on the problem of OOTA cycles, including a
brief summary of prior work.
Section~\ref{sec:OOTA and Semantic Dependencies}
motivates a shift in definition of OOTA cycle and reviews properties of
semantic dependencies, showing that these properties are a function of
executions rather than simply of the source code.
This leads into
Section~\ref{sec:What is an Execution?},
which describes properties of abstract and hardware executions, along
with the relationship between the two.

This in turn leads to
Section~\ref{sec:C++ Compilers}, which describes properties of the tool
commonly used to create hardware executions with the proper relationship
to a given set of abstract executions, namely  the compiler.
Section~\ref{sec:Hardware Dependencies, Instruction Ordering, and the
Fundamental Property}
then combines hardware dependencies and instruction ordering to arrive
at the fundamental property of semantic dependencies.

This sets the stage for
Section~\ref{sec:Real-World Constraints}
to bring real-world constraints to bear, starting with the laws of
physics, informally demonstrating that these constraints prevent
OOTA cycles containing only volatile and quasi-volatile atomics from
occurring in undefined-behavior-free C++ programs running on correctly
constructed C++ compiler-based implemen- tations that are restricted to
single-threaded analysis in our current universe.

Finally,
Section~\ref{sec:Conclusion}
presents concluding remarks.
% @@@ Future directions at some point.
